Virtual Object Layer

### 2018-01-09 RFCs/HDF5/VOL/RFC/RFC_VOL.tex
updates
Former-commit-id: 075132ce087eee5f33c09f6c78cb3bd1c8f6d8c8
minor edits, fixed LaTeX
Former-commit-id: f58e77a5ab8675ee8b4755efa0593ba44b783e8b

% Title, author, etc
\title{RFC: Virtual Object Layer}
\author{Mohamad Chaarawi}
\author{Quincey Koziol}
\date{September 16, 2014}
\rfcversion{2014-09-16.v3}
\revision{January 9, 2012}{First Draft circulated internally.}
\revision{March 9, 2012}{Update and circulate internally and externally.}
\revision{August 28, 2014}{Update with new framework changes.}
\revision{September 16, 2014}{Update with new features.}

@@

\section{Introduction}
The HDF5 data model is composed of two basic objects, groups and datasets. The data model in itself is very powerful and widely used by many applications, including High Performance Computing (HPC) applications. The HDF5 library has an internal abstraction layer called the Virtual File Layer that the library calls after it translates the HDF5 data model and object API calls to operation on blocks of bytes. The operations on those blocks of bytes are abstracted out into drivers so that we do not limit users to one underlying implementation. Several drivers are already implemented and shipped with the HDF5 library such as the POSIX (sec2) driver, MPI-I/O driver, and a memory (core) driver.

The main challenge currently in HDF5 is not in the data model itself, but in the native HDF5 single file format that has performance issues that vary widely over different platforms and has some limitations where all objects actually have to be accessible locally. Furthermore, in parallel applications, file access by several processes has to be coordinated in order to maintain strict consistency semantics to avoid file corruption. In addition, the emergence of new tiered storage architectures and object storage file systems renders the current HDF5 library in need of major implementation changes to leverage such architectures.
To address those concerns, we have been working on a new abstraction layer inside the HDF5 library called the Virtual object layer (VOL). The VOL intercepts all HDF5 API calls that could potentially access objects in the file and forwards those calls to a plugin ``object driver''. The plugins could actually store the objects in variety of ways. A plugin could, for example, have objects be distributed remotely over different platforms, provide a raw mapping of the model to the file system, or even store the data in other file formats (like native netCDF or HDF4 format). The user still gets the same data model where access is done to a single HDF5 ``container''; however the plugin object driver translates from what the user sees to how the data is actually stored. Having this abstraction layer maintains the object model of HDF5 and would allow much better usage of new object storage file systems that are targeted for Exascale systems.

@@

\section{Requirements}
The following presents a set of requirements for the VOL framework:
\begin{enumerate}
\item All HDF5 API routines that could potentially modify the HDF5 file/container should map to a VOL callback. Since the VOL plugins are implementing possible a new file format, all modifications to the container should be handed to the plugins themselves to do.
\item All HDF5 API routines that do not modify the container, but just provide memory space utilities to the application (like property lists, dataspaces, etc...) should work uniformly across all plugins. 
\item Implementing a VOL plugin should not require any change to the HDF5 library unless the plugin is an internal plugin that modifies the current native file format. 
\item Current applications, third party libraries, or tools that use the HDF5 library with the native file format should not require any change to work properly with the VOL addition.
\item Performance of applications without the VOL changes should not change. The VOL framework will introduce additional function calls to every HDF5 API call that maps to through the VOL interface, but that should be detrimental to performance.
\item Applications, third party libraries, and tools that only use the HDF5 public routines and objects should work as expected with any correct and fully implemented plugin. Performance may and probably will vary across the different plugins.
\item The VOL framework should allow plugins to be stackable on top of each other, and should allow implementation of plugins that use 2 or more different plugins. It is the plugin developer's  responsibility though to correctly implement such plugins.
\item Plugin developers should not be required to implement the full VOL functionality for a specific plugin if the intended use of that plugin is a subset of the HDF5 data model and features. For example, if a user just intends to access HDF5 datasets, it should be sufficient to develop a plugin that implement container and dataset operations in that container. Other advanced functionality of the HDF5 data model should not be required if they will not be used.
\item The HDF5 library and VOL framework should allow for new functionality or services specific to certain plugins to be added to the library and handed down to the plugin through the HDF5 library from the application.
\item Applications should be allowed to select a default plugin to use at runtime (with and environment variable possibly) or library configure time so that applications using HDF5 need not be modified to use different plugins than the native plugin.
\item VOL plugins that can be installed on a system as a shared library should be detected and loaded by the HDF5 library if the application requests their use.
\end{enumerate}

\section{Design/Architecture}
To implement a VOL interface capable of accepting any plugin a developer wishes to add, a modular design has to be put in place, similar to how the VFL interface in HDF5 is designed to accept different VFDs. The VOL is a higher level abstraction than the VFL abstraction. The following figure shows where the VOL is layered and how the data is accessed in the file:

@@

The H5Dcreate call will inform the user that a dataset with the returned ID has been created in the HDF5 container being accessed with the chosen VOL plugin. Depending on the plugin selected, the H5Dcreate function might have created a dataset object in an HDF5 file, a netCDF file, a file on a remote machine, or any way a plugin is designed to create a dataset.

@@

This would indicate that the MDS object layer plugin is to be used. All accesses to objects within the file opened using that FAPL would be routed through the MDS plugin.
In addition to plugins for the VOL class that will be added in the HDF5 library, any developer may write their own customizable plugin and tell the HDF5 library to use it at runtime. The plugin should be registered using:
{\lstsetc
\begin{lstlisting}
hid_t H5VLregister(H5VOL_class_t *cls)
\end{lstlisting}
}
The {\tt cls} parameter is a structure containing function pointers to the routines that implement all the callback routines the VOL class specifies. For more information on creating and using VOL plugins, please refer to the VOL user guide.

\subsection{Plugin Examples}
\subsubsection{Different File Format Plugins}
We mentioned earlier that a developer could design a plugin such that the file that stores data on disk could be of different format than the H5 format. This is very important to users that would like to use the HDF5 API and data model, but also desire the portability to use their files across different libraries or would like to access data from pre-existing files that are not in the H5 format.

The plugin, which understands the HDF5 data model and how the data is stored, will do the necessary ``translation'' needed for this kind of model. The following figure explains more on how this is achieved:

@@

Remote visualization is an important use case for this plugin. Large, remote datasets are very expensive to migrate to the local virtualization system. It would be faster to just enable in situ virtualization to remotely access the virtualization data using the HDF5 API. This is however not possible at the moment with the native implementation for HDF5 as the file is required to be present locally.

@@

It would be also possible to stack VOL plugins on top of each other. This notion is similar to the idea of the split VFD, where underneath the split VFD itself, two file drivers would be used, one for the file storing the metadata and another for raw data. Some stackings make sense and others would be erroneous. For example, stacking the native HDF5 plugin on top of a non-HDF5 backend plugin does not make sense and is erroneous. Figure~\ref{fig:stack} showes a stacking of the remote plugin, where data is distributed remotely, on top of the native h5 plugin, where servers that store the data at remote locations use the h5 file format.

@@

Another useful design option is to allow a mirroring plugin, where the HDF5 API calls are forwarded through a mirror plugin to two or more VOL plugins. This is an extention to the stacking feature. Figure~\ref{fig:mirror} shows an example of a VOL mirror that maps HDF5 API calls to an h5 backend plugin and an XML backend plugin:

@@

Another possible VOL plugin could be a statistics plugin that just gathers information on HDF5 API calls and records statistics associated with the number of calls to a specific API functions and corresponding parameters. This plugin would be very useful for profiling purposes. The statistics plugin would be stacked on top of another VOL plugin that actually performs the required access to the file.
\clearpage
\section{Implementation}
The VOL class structure consists of data structure(s) that contain configuration variables that are set by each plugin and a collection of function pointers that maps all HDF5 API operations that access data on disk to the plugin's corresponding implementations. Internally, each plugin is free to implement the underlying storage of objects in any way desired. The main goal is that the user still gets the same HDF5 API and data access model no matter which plugin is used. This section highlights some implementation and feature details.

@@

The VOL class includes all the higher level logical callbacks that potentially touch data (metadata or raw data) on disk. All the API routines should map to one of those callbacks. The HDF5 API is too large however to have a 1 to 1 mapping from API routines to VOL callbacks, so we introduce the notion of a generic function call that developers could use to implement certain API routines that do not map to one of the VOL callbacks. This is similar to the UNIX {\tt ioctl()} routine.

The HDF5 library currently does not support nonblocking I/O operations. However, since it is a very desirable feature, future versions of the library most probably will add nonblocking routines. In order to be able to support those API routines and not to double the number of VOL callback routines, we made all VOL callbacks nonblocking compatible, by adding an HDF5 request parameter to all VOL callbacks. If the high level operation issued by the user is nonblocking, then the request would just be forwarded to the VOL plugin and the plugin itself would be responsible to support the nonblocking behavior. If the high level operation is a blocking operation, then a non-active request is forwarded to the plugin which detects that the request is inactive and implements the operation in a blocking behavior.

One could argue that the interface can be simplified by having, for example, a common open routine for all objects (like the {\tt H5Oopen}) routine. We found however that having  a common interface would be detrimental to performance in certain plugin implementations, since it would require querying the VOL plugin twice, once for looking up the object location and once for opening the object. Having an optional callback for every object and a generic one would solve this issue. The VOL layer would check if the specific object callback is implemented in the VOL plugin, and if it is not, it would fall back to the generic object callback routine. This way we do not sacrifice performance for generality.

Another design issue comes to mind when looking at the large set of functions, and that is whether to lump all the functions together into one data structure as the VOL class, or to have a more object oriented approach, where we can either have a general class that contains all common functions, and then children of that class that contain functions specific to certain HDF5 objects, or for each object have a set of callbacks that are specific to that object. The current design chosen seems to be a good middle ground between generality and readability/maintainability.

\subsection{Internal vs. External Plugins}
Developers can implement their VOL plugin as a plugin internal to the HDF5 library or an external plugin. Internal plugins are ones that rely heavily on internal and private HDF5 functionality like the native and metadata server plugins. They are required to ship with the HDF5 library to be used by applications. External plugins on the other hand are ones that are implemented outside of the HDF5 library and applications link with to use. Those plugins do not require using private HDF5 functionality. They can however use public HDF5 features.

External plugins can be provided by the applications themselves or as a third party library, i.e. not provided by the HDF5 library nor the application. Third party plugins should be installed on the system to be used as a shared library or a DLL. HDF5 will attempt to load the plugin if the application requests its use. This is similar to Filter plugins~\cite{filters}. Refer to the VOL user guide for more detailed information on how to develop and use VOL plugins.

\subsection{{\tt H5Ocopy and H5Ocompare}}
The HDF5 API has two routines that will require extra attention to be compatible with the VOL layer, and those are the object copy and compare routines:
{\lstsetc
\begin{lstlisting}
herr_t H5Ocopy( hid_t src_loc_id, const char *src_name, hid_t dst_loc_id, const char *dst_name, hid_t ocpypl_id, hid_t lcpl_id );
htri_t H5Ocompare(hid_t loc1_id, const char *name1, hid_t lapl1,  hid_t loc2_id, const char *name2, hid_t lapl2, hid_t cmppl_id, H5O_cmp_cb_t *cb_info);
\end{lstlisting}
}

If the two objects are in containers of the same type, i.e. created using the same VOL plugin, then the VOL layer would be able to detect that and forward those calls to the corresponding implementation using a designated VOL callback. This would allow different sorts of optimizations that a VOL plugin could choose to implement in order to copy or compare two objects. On the other hand, if the objects belong to two different containers, the VOL plugin would not be able to interpret an object created by a different plugin, and so a different approach needs to be considered.

>>>>

% Title, author, etc
\title{Virtual Object Layer}
\author{Mohamad Chaarawi}
\author{Quincey Koziol}
\date{September 16, 2014}
\rfcversion{2014-09-16.v3}
\revision{January 9, 2012}{First Draft circulated internally.}
\revision{March 9, 2012}{Update and circulate internally and externally.}
\revision{August 28, 2014}{Update with new framework changes.}
\revision{September 16, 2014}{Update with new features.}

@@

\section{Introduction}
The HDF5 data model is composed of two primary objects, groups and datasets. The data model in itself is very powerful and widely used by many applications, including High Performance Computing (HPC) applications. The HDF5 library has an internal abstraction layer called the Virtual File Layer that the library calls after it translates the HDF5 data model and object API calls to operation on blocks of bytes. The operations on those blocks of bytes are abstracted out into drivers so that we do not limit users to one underlying implementation. Several drivers are already implemented and shipped with the HDF5 library such as the POSIX (sec2) driver, MPI-I/O driver, and a memory (core) driver.

The primary challenge currently in HDF5 is not in the data model itself, but in the native HDF5 single file format that has performance issues that vary widely over different platforms and has some limitations where all objects have to be accessible locally. Furthermore, in parallel applications, file access by several processes has to be coordinated to maintain strict consistency semantics to avoid file corruption. Also, the emergence of new tiered storage architectures and object storage file systems renders the current HDF5 library in need of significant implementation changes to leverage such architectures.
To address those concerns, we have been working on a new abstraction layer inside the HDF5 library called the Virtual object layer (VOL). The VOL intercepts all HDF5 API calls that could potentially access objects in the file and forwards those calls to a plugin ``object driver''. The plugins could store the objects in a variety of ways. A plugin could, for example, have objects be distributed remotely over different platforms, provide a raw mapping of the model to the file system, or even store the data in other file formats (like native netCDF or HDF4 format). The user still gets the same data model where access is done to a single HDF5 ``container''; however the plugin object driver translates from what the user sees to how the data is actually stored. Having this abstraction layer maintains the object model of HDF5 and would allow for much better use of new object storage file systems that are targeted for Exascale systems.

@@

\section{Requirements}
The following presents a set of requirements for the VOL framework:
\begin{enumerate}
\item All HDF5 API routines that could potentially modify the HDF5 file/container should map to a VOL callback. Since the VOL plugins are implementing possible a new file format, all modifications to the container should be handed to the plugins themselves to do.
\item All HDF5 API routines that do not modify the container, but just provide memory space utilities to the application (like property lists, dataspaces, etc...) should work uniformly across all plugins. 
\item Implementing a VOL plugin should not require any change to the HDF5 library unless the plugin is an internal plugin that modifies the current native file format. 
\item Current applications, third-party libraries, or tools that use the HDF5 library with the native file format should not require any change to work properly with the VOL addition.
\item Performance of applications without the VOL changes should not change. The VOL framework will introduce additional function calls to every HDF5 API call that maps to through the VOL interface, but that should be detrimental to performance.
\item Applications, third-party libraries, and tools that only use the HDF5 public routines and objects should work as expected with any correct and fully implemented plugin. Performance may and probably will vary across the different plugins.
\item The VOL framework should allow plugins to be stackable on top of each other and should allow implementation of plugins that use 2 or more different plugins. It is the plugin developer's  responsibility though to implement such plugins correctly.
\item Plugin developers should not be required to implement the full VOL functionality for a specific plugin if the intended use of that plugin is a subset of the HDF5 data model and features. For example, if a user just intends to access HDF5 datasets, it should be sufficient to develop a plugin that implements container and dataset operations in that container. Other advanced functionality of the HDF5 data model should not be required if they will not be used.
\item The HDF5 library and VOL framework should allow for new functionality or services specific to certain plugins to be added to the library and handed down to the plugin through the HDF5 library from the application.
\item Applications should be allowed to select a default plugin to use at runtime (with and environment variable possibly) or library configure time so that applications using HDF5 need not be modified to use different plugins than the native plugin.
\item VOL plugins that can be installed on a system as a shared library should be detected and loaded by the HDF5 library if the application requests their use.
\end{enumerate}
\section{Design/Architecture}
To implement a VOL interface capable of accepting any plugin a developer wishes to add, a modular design has to be put in place, similar to how the VFL interface in HDF5 is designed to accept different VFDs. The VOL is a higher level abstraction than the VFL abstraction. The following figure shows where the VOL is layered and how the data is accessed in the file:

@@

The H5Dcreate call will inform the user that a dataset with the returned ID has been created in the HDF5 container being accessed with the chosen VOL plugin. Depending on the plugin selected, the H5Dcreate function might have created a dataset object in an HDF5 file, a netCDF file, a file on a remote machine, or anyway a plugin is designed to create a dataset.

@@

This would indicate that the MDS object layer plugin is to be used. All accesses to objects within the file opened using that FAPL would be routed through the MDS plugin.
In addition to plugins for the VOL class that will be added in the HDF5 library, any developer may write their own customizable plugin and tell the HDF5 library to use it at runtime. The plugin should be registered using:
{\lstsetc
\begin{lstlisting}
hid_t H5VLregister(H5VOL_class_t *cls)
\end{lstlisting}
}
The \texttt{cls} parameter is a structure containing function pointers to the routines that implement all the callback routines the VOL class specifies. For more information on creating and using VOL plugins, please refer to the VOL user guide.

\subsection{Plugin Examples}
\subsubsection{Different File Format Plugins}
We mentioned earlier that a developer could design a plugin such that the file that stores data on disk could be of a different format than the H5 format. This is very important to users that would like to use the HDF5 API and data model, but also desire the portability to use their files across different libraries or would like to access data from pre-existing files that are not in the H5 format.

The plugin, which understands the HDF5 data model and how the data is stored, will do the necessary ``translation'' needed for this kind of model. The following figure explains more about how this is achieved:

@@

Remote visualization is an important use case for this plugin. Large, remote datasets are very expensive to migrate to the local virtualization system. It would be faster to just enable in-situ virtualization to remotely access the virtualization data using the HDF5 API. This is however not possible at the moment with the native implementation for HDF5 as the file is required to be present locally.

@@

It would also be possible to stack VOL plugins on top of each other. This notion is similar to the idea of the split VFD, where underneath the split VFD itself, two file drivers would be used, one for the file storing the metadata and another for raw data. Some stackings make sense and others would be erroneous. For example, stacking the native HDF5 plugin on top of a non-HDF5 backend plugin does not make sense and is erroneous. Figure~\ref{fig:stack} shows a stacking of the remote plugin, where data is distributed remotely, on top of the native h5 plugin, where servers that store the data at remote locations use the h5 file format.

@@

Another useful design option is to allow a mirroring plugin, where the HDF5 API calls are forwarded through a mirror plugin to two or more VOL plugins. This is an extension to the stacking feature. Figure~\ref{fig:mirror} shows an example of a VOL mirror that maps HDF5 API calls to an h5 backend plugin and an XML backend plugin:

@@

Another possible VOL plugin could be a statistics plugin that just gathers information on HDF5 API calls and records statistics associated with the number of calls to specific API functions and corresponding parameters. This plugin would be very useful for profiling purposes. The statistics plugin would be stacked on top of another VOL plugin that actually performs the required access to the file.
\clearpage
\section{Implementation}
The VOL class structure consists of data structure(s) that contain configuration variables that are set by each plugin and a collection of function pointers that maps all HDF5 API operations that access data on disk to the plugin's corresponding implementations. Internally, each plugin is free to implement the underlying storage of objects in any way desired. The main goal is that the user still gets the same HDF5 API and data access model no matter which plugin is used. This section highlights some implementation and feature details.

@@

The VOL class includes all the higher level logical callbacks that potentially touch data (metadata or raw data) on disk. All the API routines should map to one of those callbacks. The HDF5 API is too large however to have a 1 to 1 mapping from API routines to VOL callbacks, so we introduce the notion of a generic function call that developers could use to implement certain API routines that do not map to one of the VOL callbacks. This is similar to the UNIX \texttt{ioctl()} routine.

The HDF5 library currently does not support nonblocking I/O operations. However, since it is a very desirable feature, future versions of the library most probably will add nonblocking routines. In order to be able to support those API routines and not to double the number of VOL callback routines, we made all VOL callbacks nonblocking compatible, by adding an HDF5 request parameter to all VOL callbacks. If the high-level operation issued by the user is nonblocking, then the request would just be forwarded to the VOL plugin and the plugin itself would be responsible to support the nonblocking behavior. If the high level operation is a blocking operation, then a non-active request is forwarded to the plugin which detects that the request is inactive and implements the operation in a blocking behavior.

One could argue that the interface can be simplified by having, for example, a common open routine for all objects (like the \texttt{H5Oopen}) routine. We found however that having  a common interface would be detrimental to performance in certain plugin implementations, since it would require querying the VOL plugin twice, once for looking up the object location and once for opening the object. Having an optional callback for every object and a generic one would solve this issue. The VOL layer would check if the specific object callback is implemented in the VOL plugin, and if it is not, it will fall back to the generic object callback routine. This way we do not sacrifice performance for generality.

Another design issue comes to mind when looking at the large set of functions, and that is whether to lump all the functions together into one data structure as the VOL class, or to have a more object-oriented approach, where we can either have a general class that contains all common functions, and then children of that class that contain functions specific to certain HDF5 objects, or for each object have a set of callbacks that are specific to that object. The current design chosen seems to be a good middle ground between generality and readability/maintainability.

\subsection{Internal vs. External Plugins}
Developers can implement their VOL plugin as a plugin internal to the HDF5 library or an external plugin. Internal plugins are ones that rely heavily on internal and private HDF5 functionality like the native and metadata server plugins. They are required to ship with the HDF5 library to be used by applications. External plugins, on the other hand, are ones that are implemented outside of the HDF5 library and applications link with to use. Those plugins do not require using private HDF5 functionality. They can, however, use public HDF5 features.

External plugins can be provided by the applications themselves or as a third party library, i.e., not provided by the HDF5 library nor the application. Third party plugins should be installed on the system to be used as a shared library or a DLL. HDF5 will attempt to load the plugin if the application requests its use. This is similar to Filter plugins~\cite{filters}. Refer to the VOL user guide for more detailed information on how to develop and use VOL plugins.

\subsection{\texttt {H5Ocopy and H5Ocompare}}
The HDF5 API has two routines that will require extra attention to be compatible with the VOL layer, and those are the object copy and compare routines:
{\lstsetc
\begin{lstlisting}
herr_t H5Ocopy( hid_t src_loc_id, const char *src_name, hid_t dst_loc_id, const char *dst_name, hid_t ocpypl_id, hid_t lcpl_id );
htri_t H5Ocompare(hid_t loc1_id, const char *name1, hid_t lapl1,  hid_t loc2_id, const char *name2, hid_t lapl2, hid_t cmppl_id, H5O_cmp_cb_t *cb_info);
\end{lstlisting}
}

If the two objects are in containers of the same type, i.e., created using the same VOL plugin, then the VOL layer would be able to detect that and forward those calls to the corresponding implementation using a designated VOL callback. This would allow different sorts of optimizations that a VOL plugin could choose to implement in order to copy or compare two objects. On the other hand, if the objects belong to two different containers, the VOL plugin would not be able to interpret an object created by a different plugin, and so a different approach needs to be considered.
